Interactive Visualization of Hierarchically Structured Data

### 2017-03-18 treelapse.Rmd
Replace microbe with bacteria

Our key contribution is the adaptation of the visualization principles
of focus-plus-context and linking to the study of tree-structured data
[@buja1996interactive, @becker1987brushing]. Our motivating
application is to the analysis of microbial time series, where an
evolutionary tree relating microbes is available a priori. However, we
have identified common problem types where, if a tree is not directly
available, it can be constructed from data and then studied using our
techniques.

We have implemented our visualizations in D3, but encapsulated in an R package,
called treelapse, to facilitate rapid turnover from data preparation and
modeling to interactive exploration, and vice versa. Our code is open-source,
and available at \url{xxxxxxxxx}\footnote{Link removed in blinded manuscript,
but available to editor.}. We hope this package encourages data analysts to work
at the border between data modeling and visualization, and more generally
empowers a wider audience to apply less widely known, but powerful,
visualization ideas.

@@

A microbiome is a community of microbes living in given environments,
for example, ocean water or the human gut [@human2012structure,
@human2012structure, @cho2012human]. Progress in the field has been
rapidly accelerated by the advent of genomic technologies, which
enable detailed quantification of microbial ecological structure and
its influence in human and environmental health. Being concerned with
both microbial community structure and human health, the field exists
at the border between ecology and medicine; consequently, papers in
the area often apply a blend of exploratory data analysis and formal
statistical inference.

The two essential microbiome analysis problems that motivated our work
are the tree-structured differential abundance and differential
dynamics problems. In the differential abundance problem, we attempt
to compare the abundances of individual microbes across experimental
conditions -- for example, treatment vs. control or healthy
vs. diseased. This is the microbiome analog of differential expression
analysis in genomics [@anders2010differential]. We prepend the
description "tree-structured" because, in practice, researchers
generate interpretations about intermediate taxonomic orders of
microbes -- it is more interesting to discover novel behavior
taxonomic levels between high-order phyla and low-level
species. Hence, we frame the tree-structured differential abundance
problem as the question of identifying the largest taxonomic subtree
whose associated microbes are differentially abundant.

In the tree structured microbial dynamics problem, the goal is to
describe changes in microbial abundances in an environment over
time. As in the differential abundance problem, it is useful if these
descriptions can be given at the highest subtree at which the pattern
appears. Specific questions of interest often have an ecological
flavor. For example, researchers are often interested in understanding
how microbial populations respond to sudden or gradual environmental
changes or how microbes fill, drop out from, or compete for
environmental niches. Medically, these questions are important for
illuminating the impact of antibiotic time courses or diet changes,
for example.

@@

To unify the tree-structured differential abundance and microbial
dynamic problems, we identify the data with a collection of random
variables indexed by nodes in a prespecified tree structure. In the
differential abundance problem, each random variable lives in
$\mathbb{R}^{G}$ where $G$ is the number of groups being
compared. Each coordinate represents the abundance for that group, and
a node exhibits differential abundance when the coordinates are drawn
from different distributions. On the other hand, in the microbial
dynamics problem, each random variable is a time series, living in
$\mathbb{R}^{T}$.

@@

Our third display is directed at the microbial dynamics
question. Here, two panels are arranged one over the other; one
displays all time series, while the other displays all tree nodes,
with node sizes reflecting the time series value at that node. For
this reason, we call the display, Timebox Trees. In the time series
panel, we have directly implemented the timeboxes idea. We then link
the panels: when a set of series is highlighted by the timeboxes, the
associated tree nodes are also highlighted. For example, timeboxes can
be used to focus on a set of series with specific shape -- increased
abundance after an ecological shock, for example -- and identify along
what subtrees this pattern is present. To further focus on specific
elements, a pan-zoom scented widget is provided
[@willett2007scented]. The widget is a miniature version of the full
time series panel, equipped with a single brush whose extent specifies
the limits in the main time series panel. As in the DOI trees and
sankeys, a search bar can be used to highlight those series of
interest a priori.

The final display currently implemented in the package is the natural
converse of the timebox trees display. Rather than defining visual
queries in terms of time series, it defines queries using nodes in the
tree. For this reason, we call the display Treeboxes. Rather than
focusing on the intersection of brushes, as in timebox trees, we focus
on the union of brushed over nodes. This allows us to highlight series
associated with nodes on distant subtrees. This display is also suited
for the microbial dynamics problem. For example, by highlighting all
nodes at one taxonomic level in the tree, we can easily summarize the
time series pattern for all the taxa at that level. Alternatively,
focusing on all the children below a single node makes it possible to
see how much correlation and competition there is between
taxonomically similar microbes. As in the timebox trees display, a
search box and pan-zoom scented widget are provided.

@@

### Microbial Dynamics of Antibiotics Time Courses ###

[@dethlefsen2008pervasive] investigates the effect of antibiotics on
microbial community composition from an ecological perspective. The
study tracks the microbiome of three patients across ten months, with
two five-day antibiotic time courses separated by 6 months. Discerning
the variation in resilience across microbes is important, considering
the the role of microbes in health and not just disease.

We approach the data using the linked time and treebox views, after
first filtering low variance taxa and taking an $\text{asinh}$
transformation. An initial view, Figure \@ref(fig:antibioticoverview),
reveals two dramatic drops in the overall microbial abundance time
series during the antibiotics time courses. Two more subtle effects
are also suggested from this view,

@@

Alternatively, we can summarize each node by the average across its
descendants -- this brings attention to individual microbes that may
be underlying some of the broader taxonomic patterns we have noted
when studying the subtree sums. For example, in Figure
\@ref(fig:ruminococcus), we highlight all families below order
Ruminococcus, suggesting that the decrease due to antibiotics occurs
uniformly across almost all families. A point that was not evident in
the earlier sum-across-descendants view is that, after the second
treatment of antibiotics, a few of the Ruminoccocus families recover
more rapidly than the rest, for example the Unc095d3 (highlighted in
red) are only briefly affected. In contrast, most families seem to
recover in unison after the first treatment.

@@

Further, note that in this subtree averages view, the tree display has
changed. This is because, at each branching point, we place the node
with larger average value on the left. This places the
Verrucomicrobiae at the far left, which seem to have large abundance
over many time points. This phylum had been previously obfuscated --
because there are not many leaves associated with this phylum, the sum
was small. Interestingly, the abundance of these microbes seems to
*increase* after the first antibiotics treatment.  Be cautious,
however, that the average over only a few Verrucomicrobiae species
will be a more variable estimate than the averages over the more
prevalent phyla.

@@

### Differential Microbial Abundance and Preterm Births ###

The study [@digiulio2015temporal] tracked the abundance of microbes in
the vaginal microbiome during pregnancy in an effort to study
relationships between microbial community composition and preterm
birth. Ideally, it would be possible to develop clear microbial
signatures associated with preterm births.

@@

Therefore, our focus here is on the differential abundance question,
rather than dynamics.  We would like to provide visual representations
of differential abundance across CSTs and also between preterm and
non-preterm births. [@digiulio2015temporal] interpreted the CSTs using
a heatmap, with microbes ordered according to a hierarchical
clustering.  By using the DOI sankey instead, we can interpret the
CSTs in their taxonomic context and at multiple scales of taxonomic
resolution. Further, while [@digiulio2015temporal] focused on
identifying associations between preterm births and CSTs -- presumably
because testing individual microbes loses power -- we can compare
microbe abundances between preterm and non-preterm samples along
subtrees, without requiring CSTs as an intermediary.

In Figure \@ref(fig:pretermcsts), we compare the 5 CSTs according to
their values along the subtree. Specifically, we took the average of
all samples within each CST to define values at the (microbe-level)
leaf nodes, and then aggregated the averages up to the root. It is
immediately clear that samples from CST 4 have much more taxonomic
diversity. Further, focusing on the Lactobacillaceae family, we note
that the differential abundance of these microbes distinguishes the
remaining CSTs, see Figure \@ref(fig:pretermcstslacto).

@@

```{r pretermpreterm, fig.cap = "Samples with high levels of phyla other than Firmicutes appear to be related to preterm births."}
include_graphics("figure/preterm_preterm.png")
```

@@

Each of the timebox tree and treebox examples presented so far have focused on
data with a clear time component. We note however that these techniques could
alternatively be applied to high-dimensional data, via the use of parallel
coordinates [@inselberg1991parallel]. The usual parallel coordinates challenges
remain, namely selecting scales for and an ordering across the different
coordinates, but the linking and focus-plus-context ideas can still be employed
in this setting. Here we provide an implementation of this idea on a dataset
comparing microbiomes across various ecological environments
[@caporaso2011global], which is publicly accessible through phyloseq
[@mcmurdie2013phyloseq].

@@

We hierarchically cluster these 26 samples based on the 500 most abundant taxa,
using complete linkage on the UniFrac distance. Figure \@ref(gptimebox) displays
the resulting hierarchy together with a parallel coordinates view of the
$\text{asinh}$ transformed taxa.

```{r gptimebox, fig.cap = "An application to the Global Patterns demonstrates how linking in treelapse can be applied to combine hierarchical clustering and parallel coordinates views."}
include_graphics(c("figure/gp_cluster1.png", "figure/gp_cluster2.png"))
```

In Figure \@ref(fig:gptimebox), we compare two subclusters from the hierarchical
clustering tree, after zooming to a few of the microbes that distinguish between
the clusters. Upon revisiting the original data, it becomes clear that the
samples highlighted on the left come from freshwater samples, while those on the
right come from soil and skin, and looking up taxonomic groups associated with
the distinguishing microbes confirms this. For example, many of the microbes
with high abundances in the left figure come from order Oceanospirillales.

@@

In summary, we have built an easily accessible R package for
visualization techniques in a very specific methodology problem –
analysis of differential abundance and dynamics in hierarchically
structured data – that appears in a variety of application domains. We
have leveraged a link between R and D3 [@vaidyanathan2014htmlwidgets]
to create exploratory visualizations while still deep in data
analysis; in this way our work is a departure from the culture of
polished, journalistic visualizations prioritized by the D3
community. Finally, we have given a series of examples to demonstrate
how the general visualization techniques of focus-plus-context and
linked brushing can be practically incorporated into a range of
practical analysis workflows, from studying the impact of microbes on
human health to better allocating units in commuter bikesharing
systems.

>>>>

Our key contribution is the adaptation of the visualization principles
of focus-plus-context and linking to the study of tree-structured data
[@buja1996interactive, @becker1987brushing]. Our motivating
application is to the analysis of bacterial time series, where an
evolutionary tree relating bacteria is available a priori. However, we
have identified common problem types where, if a tree is not directly
available, it can be constructed from data and then studied using our
techniques.

We have implemented our visualizations in D3, but encapsulated in an R package,
called treelapse, to facilitate rapid turnover from data preparation and
modeling to interactive exploration, and vice versa. Our code is open-source,
and available at \url{https://github.com/krisrs1128/treelapse}\footnote{Link removed in blinded manuscript,
but available to editor.}. We hope this package encourages data analysts to work
at the border between data modeling and visualization, and more generally
empowers a wider audience to apply less widely known, but powerful,
visualization ideas.

@@

A microbiome is a community of bacteria living in given environments,
for example, ocean water or the human gut [@human2012structure,
@human2012structure, @cho2012human]. Progress in the field has been
rapidly accelerated by the advent of genomic technologies, which
enable detailed quantification of bacterial ecological structure and
its influence in human and environmental health. Being concerned with
both bacterial community structure and human health, the field exists
at the border between ecology and medicine; consequently, papers in
the area often apply a blend of exploratory data analysis and formal
statistical inference.

The two essential microbiome analysis problems that motivated our work
are the tree-structured differential abundance and differential
dynamics problems. In the differential abundance problem, we attempt
to compare the abundances of individual bacteria across experimental
conditions -- for example, treatment vs. control or healthy
vs. diseased. This is the microbiome analog of differential expression
analysis in genomics [@anders2010differential]. We prepend the
description "tree-structured" because, in practice, researchers generate
interpretations about intermediate taxonomic orders -- it is more interesting to
discover novel behavior taxonomic levels between high-order phyla and low-level
species. Hence, we frame the tree-structured differential abundance problem as
the question of identifying the largest taxonomic subtree whose associated
bacteria are differentially abundant.

In the tree structured bacterial dynamics problem, the goal is to
describe changes in bacterial abundances in an environment over
time. As in the differential abundance problem, it is useful if these
descriptions can be given at the highest subtree at which the pattern
appears. Specific questions of interest often have an ecological
flavor. For example, researchers are often interested in understanding
how bacterial populations respond to sudden or gradual environmental
changes or how species fill, drop out from, or compete for
environmental niches. Medically, these questions are important for
illuminating the impact of antibiotic time courses or diet changes,
for example.

@@

To unify the tree-structured differential abundance and bacterial
dynamics problems, we identify the data with a collection of random
variables indexed by nodes in a prespecified tree structure. In the
differential abundance problem, each random variable lives in
$\mathbb{R}^{G}$ where $G$ is the number of groups being
compared. Each coordinate represents the abundance for that group, and
a node exhibits differential abundance when the coordinates are drawn
from different distributions. On the other hand, in the bacterial
dynamics problem, each random variable is a time series, living in
$\mathbb{R}^{T}$.

@@

Our third display is directed at the bacterial dynamics
question. Here, two panels are arranged one over the other; one
displays all time series, while the other displays all tree nodes,
with node sizes reflecting the time series value at that node. For
this reason, we call the display, Timebox Trees. In the time series
panel, we have directly implemented the timeboxes idea. We then link
the panels: when a set of series is highlighted by the timeboxes, the
associated tree nodes are also highlighted. For example, timeboxes can
be used to focus on a set of series with specific shape -- increased
abundance after an ecological shock, for example -- and identify along
what subtrees this pattern is present. To further focus on specific
elements, a pan-zoom scented widget is provided
[@willett2007scented]. The widget is a miniature version of the full
time series panel, equipped with a single brush whose extent specifies
the limits in the main time series panel. As in the DOI trees and
sankeys, a search bar can be used to highlight those series of
interest a priori.

The final display currently implemented in the package is the natural
converse of the timebox trees display. Rather than defining visual
queries in terms of time series, it defines queries using nodes in the
tree. For this reason, we call the display Treeboxes. Rather than
focusing on the intersection of brushes, as in timebox trees, we focus
on the union of brushed over nodes. This allows us to highlight series
associated with nodes on distant subtrees. This display is also suited
for the bacterial dynamics problem. For example, by highlighting all
nodes at one taxonomic level in the tree, we can easily summarize the
time series pattern for all the taxa at that level. Alternatively,
focusing on all the children below a single node makes it possible to
see how much correlation and competition there is between
taxonomically similar bacteria. As in the timebox trees display, a
search box and pan-zoom scented widget are provided.

@@

### Bacterial Dynamics of Antibiotics Time Courses ###

[@dethlefsen2008pervasive] investigates the effect of antibiotics on
bacterial community composition from an ecological perspective. The
study tracks the microbiome of three patients across ten months, with
two five-day antibiotic time courses separated by 6 months. Discerning
the variation in resilience across bacteria is important, considering
the the role of bacteria in health and not just disease.

We approach the data using the linked time and treebox views, after
first filtering low variance taxa and taking an $\text{asinh}$
transformation. An initial view, Figure \@ref(fig:antibioticoverview),
reveals two dramatic drops in the overall bacterial abundance time
series during the antibiotics time courses. Two more subtle effects
are also suggested from this view,

@@

Alternatively, we can summarize each node by the average across its
descendants -- this brings attention to individual bacteria that may
be underlying some of the broader taxonomic patterns we have noted
when studying the subtree sums. For example, in Figure
\@ref(fig:ruminococcus), we highlight all families below order
Ruminococcus, suggesting that the decrease due to antibiotics occurs
uniformly across almost all families. A point that was not evident in
the earlier sum-across-descendants view is that, after the second
treatment of antibiotics, a few of the Ruminoccocus families recover
more rapidly than the rest, for example the Unc095d3 (highlighted in
red) are only briefly affected. In contrast, most families seem to
recover in unison after the first treatment.

@@

Further, note that in this subtree averages view, the tree display has
changed. This is because, at each branching point, we place the node
with larger average value on the left. This places the
Verrucomicrobiae at the far left, which seem to have large abundance
over many time points. This phylum had been previously obfuscated --
because there are not many leaves associated with this phylum, the sum
was small. Interestingly, the abundance of these bacteria seems to
*increase* after the first antibiotics treatment.  Be cautious,
however, that the average over only a few Verrucomicrobiae species
will be a more variable estimate than the averages over the more
prevalent phyla.

@@

### Differential Bacterial Abundance and Preterm Births ###

The study [@digiulio2015temporal] tracked the abundance of bacteria in
the vaginal microbiome during pregnancy in an effort to study
relationships between bacterial community composition and preterm
birth. Ideally, it would be possible to develop clear bacterial
signatures associated with preterm births.

@@

Therefore, our focus here is on the differential abundance question,
rather than dynamics.  We would like to provide visual representations
of differential abundance across CSTs and also between preterm and
non-preterm births. [@digiulio2015temporal] interpreted the CSTs using
a heatmap, with bacteria ordered according to a hierarchical
clustering.  By using the DOI sankey instead, we can interpret the
CSTs in their taxonomic context and at multiple scales of taxonomic
resolution. Further, while [@digiulio2015temporal] focused on
identifying associations between preterm births and CSTs -- presumably
because testing individual bacteria loses power -- we can compare
bacterial abundances between preterm and non-preterm samples along
subtrees, without requiring CSTs as an intermediary.

In Figure \@ref(fig:pretermcsts), we compare the 5 CSTs according to
their values along the subtree. Specifically, we took the average of
all samples within each CST to define values at the (species-level)
leaf nodes, and then aggregated the averages up to the root. It is
immediately clear that samples from CST 4 have much more taxonomic
diversity. Further, focusing on the Lactobacillaceae family, we note
that the differential abundance of these bacteria distinguishes the
remaining CSTs, see Figure \@ref(fig:pretermcstslacto).

@@

```{r pretermpreterm, out.width = "300px", fig.cap = "Samples with high levels of phyla other than Firmicutes appear to be related to preterm births."}
include_graphics("figure/preterm_preterm.png")
```

@@

Each of the timebox tree and treebox examples presented so far have focused on
data with a clear time component. We note however that these techniques could
alternatively be applied to high-dimensional data, via the use of parallel
coordinates [@inselberg1991parallel]. The usual parallel coordinates challenges
remain, namely selecting scales for and an ordering across the different
coordinates, but the linking and focus-plus-context ideas can still be employed
in this setting. Here we provide an implementation of this idea on a dataset
comparing microbiomes across various ecological environments
[@caporaso2011global], which is publicly accessible through the phyloseq R
package [@mcmurdie2013phyloseq].

@@

We hierarchically cluster these 26 samples based on the 500 most abundant taxa,
using complete linkage on the UniFrac distance. Figure \@ref(fig:gptimebox) displays
the resulting hierarchy together with a parallel coordinates view of the
$\text{asinh}$ transformed taxa.

```{r gptimebox, out.width = "200px", fig.show = "hold", fig.cap = "An application to the Global Patterns demonstrates how linking in treelapse can be applied to combine hierarchical clustering and parallel coordinates views."}
include_graphics(c("figure/gp_cluster1.png", "figure/gp_cluster2.png"))
```

In Figure \@ref(fig:gptimebox), we compare two subclusters from the hierarchical
clustering tree, after zooming to a few of the bacteria that distinguish between
the clusters. Upon revisiting the original data, it becomes clear that the
samples highlighted on the left come from freshwater samples, while those on the
right come from soil and skin, and looking up taxonomic groups associated with
the distinguishing bacteria confirms this. For example, many of the species
with high abundances in the left figure come from order Oceanospirillales.

@@

In summary, we have built an easily accessible R package for
visualization techniques in a very specific methodology problem –
analysis of differential abundance and dynamics in hierarchically
structured data – that appears in a variety of application domains. We
have leveraged a link between R and D3 [@vaidyanathan2014htmlwidgets]
to create exploratory visualizations while still deep in data
analysis; in this way our work is a departure from the culture of
polished, journalistic visualizations prioritized by the D3
community. Finally, we have given a series of examples to demonstrate
how the general visualization techniques of focus-plus-context and
linked brushing can be practically incorporated into a range of
practical analysis workflows, from studying the impact of bacteria on
human health to better allocating units in commuter bikesharing
systems.
